A meta-analysis of fungicide effects on mungbean powdery mildew

### 2021-12-28 04_Severity_MetaAnalysis.Rmd
Revert major changes to analysis, leave incorrect analysis in place

This model will be similar to the grain yield meta-analysis and use the main variables:
   - *Powdery mildew severity*, the response variable (1-9 scale)
   - *Trial*, which resolves combinations of categorical random variables: 
      i) year
      ii) location
      iii) row spacing
      iv) cultivar
   - *spray_management*, a moderator to evaluate the difference in effect size attributed to fungicide application timing.  
   - *id*, random variable indicating each treatment is independent

@@

### Define Trial

```{r define_trial}
PM_dat_D <-
   PM_dat_D %>%
   mutate(trial = paste(trial_ref,
                        year,
                        location,
                        host_genotype,
                        row_spacing,
                        sep = "_")) 
```

@@

Because the data were taken on an ordinal scale, we need to transform this to a numeric value to perform the analysis, _i.e._

> For ordinal measurements (e.g., disease rating on a 0 to 3 scale designating the ordinal degree of symptoms), nonparametric methods based on ranks should generally be used for hypothesis testing. See Shah and Madden (2004) for recommendations on the treatment of ordinal data. As an alternative, parametric proportional-odds models may be used for ordinal data, if there are sufficiently large numbers of observations for each experimental unit. If ordinal rating scores are used for ranges of disease severity (e.g., Horsfall-Barratt [where, for instance, a “2” corresponds to a severity range from 3 to 6%, and a “3” corresponds to a range of 6 to 12%]), each score should be converted to the midpoint of the corresponding disease severity range prior to use of a parametric analytical technique.

- From <https://apsjournals.apsnet.org/page/authorinformation>

@@

```{r ord-to-perc}
# Percent severity conversion
# define ordinal conversion from scale to proportion
s1 <- 0
s2 <- 0.33 * 0.5
s3 <- 0.5 * 0.87
s4 <- 0.66 * 0.75
s5 <- 0.66 * 0.87
s6 <- 0.66 * 1
s7 <- 0.75 * 1
s8 <- 1 * 0.87
s9 <- 1
x <- c(s1, s2, s3, s4, s5, s6, s7, s8, s9)
y <- c(1:9)
a = data.frame(x, y)
plot(a)
```

@@

```{r fit-gam}
# fit a curve using a gam with spline value 0.01
g1 <- gam(x ~ s(y, spar = 0.1), family = gaussian)

# View plot of the GAM
plot(x = predict(g1,
                 data.frame(y = seq(1, 9, by = 0.1))),
     y = seq(1, 9, by = 0.1))
points(x, y, col = "red", pch = 16)

@@

## Disease severity spray schedule meta-analysis

First let's look at the results of the meta-analysis without transformation

```{r Metafor-Sev-analysis, cache=FALSE}
PMsev_mv <- rma.mv(
   yi = PM_final_severity,
   vi,
   mods = ~ spray_management,
   method = "ML",
   random = list(~ spray_management | trial, ~ 1 | id),
   struct = "UN",
   #control = list(optimizer = "optim"),
   data = PM_dat_D
)
summary(PMsev_mv)
plot(rstandard(PMsev_mv)$z, ylim = c(-3, 3), pch = 19)
```

@@

```{r Metafor-Sev-analysisSQ, cache=FALSE}
# create transformations
PM_dat_D$yi_square <- PM_dat_D$PM_final_severity ^ 2
PM_dat_D$vi_square <- (PM_dat_D$vi + 1) ^ 2

PMsev_mvSQ <- rma.mv(
   yi = yi_square,
   V = vi_square,
   mods = ~ spray_management,
   method = "ML",
   random = list(~ spray_management | trial, ~ 1 | id),
   struct = "UN",
   control = list(optimizer = "optim"),
   data = PM_dat_D
)

summary(PMsev_mvSQ)
plot(rstandard(PMsev_mvSQ)$z,
     ylim = c(-3, 3),
     pch = 19)
```

This meta-analysis is more in-line with estimates from the yield meta-analysis.
`Early` sprays are not effective at reducing powdery mildew severity and two or more sprays had higher efficacy than single sprays.
Recommended sprays were the most effective spray timing at reducing disease severity.

@@

```{r Metafor-Sev-analysis_log, cache=FALSE}
# remove controls for response ratio meta-analysis
PMsev_mv_log <- rma.mv(
   yi = yi_log,
   V = yi_log,
   mods = ~ spray_management,
   method = "ML",
   random = list(~ spray_management | trial, ~ 1 | id),
   struct = "UN",
   control = list(optimizer = "optim"),
   data = PM_dat_D
)

@@

```{r Metafor-Sev-analysis_logitP, cache=FALSE}
# remove controls for response ratio meta-analysis
PMsev_mvPer <- rma.mv(
   yi = yi_logit,
   V = yi_logit,
   mods = ~ spray_management,
   method = "ML",
   random = list(~ spray_management | trial, ~ 1 | id),
   struct = "UN",
   control = list(optimizer = "optim"),
   data = PM_dat_D
)

@@

```{r Sensitivity_logitP, cache=FALSE}
s_analysis <-
   lapply(unique(PM_dat_D$trial_ref), function(Trial) {
      dat <- filter(PM_dat_D, trial_ref != Trial)
      # remove controls for response ratio meta-analysis
      rma.mv(
         yi = yi_logit,
         V = yi_logit,
         mods = ~ spray_management,
         method = "ML",
         random = list( ~ spray_management | trial, ~ 1 | id),
         struct = "UN",
         control = list(optimizer = "optim"),
         data = dat
      )

@@

```{r}
W <- diag(1/(PM_dat_D$vi_logit+0.0001))
X <- model.matrix(PMsev_mvPer)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W
100 * sum(PMsev_mvPer$sigma2) / (sum(PMsev_mvPer$sigma2) + (PMsev_mvPer$k-PMsev_mvPer$p)/sum(diag(P)))

100 * PMsev_mvPer$sigma2 / (sum(PMsev_mvPer$sigma2) + (PMsev_mvPer$k-PMsev_mvPer$p)/sum(diag(P)))
```

>>>>

This model will be similar to the grain yield meta-analysis and use the main variables:

   - *Powdery mildew severity*, the response variable (1-9 scale)

   - *Trial*, which resolves combinations of categorical random variables: 
      i) year  
      ii) location  
      iii) row spacing  
      iv) cultivar

   - *spray_management*, a moderator to evaluate the difference in effect size attributed to fungicide application timing.

   - *id*, random variable indicating each treatment is independent

@@

<br>  
<br>

### Define Trial
```{r define_trial}
PM_dat_D <-
   PM_dat_D %>%
   mutate(trial = paste(trial_ref,
                        year,
                        location,
                        host_genotype,
                        row_spacing,
                        sep = "_")) 
```

<!--- I would remove from this line, 77 to line 96 as we shouldn't be promoting the idea of analysing an ordinal scale directly, just go straight to the severity conversion step and describe it and why it's done (i.e., you can't analyse an ordinal scale like this, though I found a cool presentation on how to analyse an ordinal scale using metanalysis for future reference, https://methods.cochrane.org/sites/methods.cochrane.org.statistics/files/public/uploads/SMG_training_course_cardiff/2010_SMG_training_cardiff_day1_session2_lewis.pdf) -->

### Inspect PM severity for normality
Given the meta-analysis assumes powdery mildew severity observations to be a univariate normal distribution $Y_i = N(\mu , \sigma)$, with a mean $\mu$ and standard deviation $\sigma$. We need to inspect severity for a normality.
```{r sev_hist}
hist(PM_dat_D$PM_final_severity)
summary(PM_dat_D$PM_final_severity)
```
We can see here that powdery mildew severity observations are skewed towards the top of the scale, eight and nine in this case.

We can improve the normality by transforming the effect size, disease severity, by squaring.
```{r}
PM_dat_D$yi_square <- PM_dat_D$PM_final_severity ^2
PM_dat_D$vi_square <- (PM_dat_D$vi +1)^2
# Note: Here I added 1 to the variance before it was squared. This is because squaring numbers below one makes the number smaller or remain zero. This does not change the meta-analysis mean estimates but does increase heterogeneity and tao._

hist(PM_dat_D$yi_square)
summary(PM_dat_D$yi_square)
```

Another transformation which might be more suitable for a discrete response is either a `log()` or `logit()` transformation. 
However, first we should convert the scale from 1 - 9 to between 0 and 1.

@@

Let us also transform the severity onto a percentage scale so we can compare the results of meta-analyses with no transformation, squared severity, and percent severity.

@@

```{r}
# Percent severity conversion
# define ordinal conversion from scale to proportion
s1 <- 0
s2 <- 0.33 * 0.5
s3 <- 0.5 * 0.87
s4 <- 0.66 * 0.75
s5 <- 0.66 * 0.87
s6 <- 0.66 * 1
s7 <- 0.75 * 1
s8 <- 1 * 0.87
s9 <- 1
x <- c(s1, s2, s3, s4, s5, s6, s7, s8, s9)
y <- c(1:9)
a = data.frame(x, y)
plot(a)
```

@@

```{r}
# fit a curve using a gam with spline value 0.01
g1 <- gam(x ~ s(y, spar = 0.1), family = gaussian)

# View plot of gam
plot(x = predict(g1,
                 data.frame(y = seq(1, 9, by = 0.1))), y = seq(1, 9, by = 0.1))
points(x, y, col = "red", pch = 16)

@@

## Disease severity spray schedule meta-analysis  
First let's look at the results of the meta-analysis without transformation
```{r Metafor-Sev-analysis, cache=TRUE}
PMsev_mv <- rma.mv(
   yi = PM_final_severity,
   vi,
   mods = ~ spray_management,
   method = "ML",
   random = list(~ spray_management | trial, ~ 1 | id),
   struct = "UN",
   #control = list(optimizer = "optim"),
   data = PM_dat_D
)
summary(PMsev_mv)
plot(rstandard(PMsev_mv)$z, ylim = c(-3, 3), pch = 19)
```

@@

```{r Metafor-Sev-analysisSQ, cache=TRUE}
PMsev_mvSQ <- rma.mv(
   yi = yi_square,
   V = vi_square,
   mods = ~ spray_management,
   method = "ML",
   random = list(~ spray_management | trial, ~ 1 | id),
   struct = "UN",
   control = list(optimizer = "optim"),
   data = PM_dat_D
)

summary(PMsev_mvSQ)
plot(rstandard(PMsev_mvSQ)$z,
     ylim = c(-3, 3),
     pch = 19)
```
This meta-analysis is more in-line with estimates from the yield meta-analysis.
`Early` sprays are not effective at reducing powdery mildew severity and two or more sprays had higher efficacy than single sprays.
Recommended sprays were the most effective spray timing at reducing disease severity.

@@

```{r Metafor-Sev-analysis_log, cache=TRUE}
# remove controls for response ratio meta-analysis
PMsev_mv_log <- rma.mv(
   yi = yi_log,
   V = yi_log,
   mods = ~ spray_management,
   method = "ML",
   random = list(~ spray_management | trial, ~ 1 | id),
   struct = "UN",
   control = list(optimizer = "optim"),
   data = PM_dat_D
)

@@

```{r Metafor-Sev-analysis_logitP, cache=TRUE}
# remove controls for response ratio meta-analysis
PMsev_mvPer <- rma.mv(
   yi = yi_logit,
   V = yi_logit,
   mods = ~ spray_management,
   method = "ML",
   random = list(~ spray_management | trial, ~ 1 | id),
   struct = "UN",
   control = list(optimizer = "optim"),
   data = PM_dat_D
)

@@

```{r Sensitivity_logitP, cache=TRUE}
s_analysis <-
   lapply(unique(PM_dat_D$trial_ref), function(Trial) {
      dat <- filter(PM_dat_D, trial_ref != Trial)
      # remove controls for response ratio meta-analysis
      rma.mv(
         yi = yi_logit,
         V = yi_logit,
         mods = ~ spray_management,
         method = "ML",
         random = list( ~ spray_management | trial, ~ 1 | id),
         struct = "UN",
         control = list(optimizer = "optim"),
         data = dat
      )

@@

```{r}
W <- diag(1 / (PM_dat_D$vi_logit + 0.0001))
X <- model.matrix(PMsev_mvPer)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W
100 * sum(PMsev_mvPer$sigma2) / (sum(PMsev_mvPer$sigma2) + (PMsev_mvPer$k -
                                                               PMsev_mvPer$p) / sum(diag(P)))

100 * PMsev_mvPer$sigma2 / (sum(PMsev_mvPer$sigma2) + (PMsev_mvPer$k - PMsev_mvPer$p) /
                               sum(diag(P)))
```
